### 2025-05-07 Chapters/Chapter4.tex
temp

\begin{table}[h]
  \caption[Model results for different sets of parameters.]{Models defined by a selection of parameter sets, with increasing complexity. The table shows decreasing MAPE as more variables are added.}
    \label{table:setsOfParams}
    \centering
    \begin{tabular}{lc}
        \toprule
        Model variables & MAPE\\
        \midrule
        $[\,]$ Baseline & 23.9\%\\
        $[ws_{15}]$ Baseline & 17.2\%\\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd]$ & 13.8\% \\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ & 13.7\%\\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ + DEM & 12.0\%\\
        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, \text{altitude}, twd, N^2, Ri]$ & 12.0\% \\
        \bottomrule
    \end{tabular}
\end{table}

\section{Modeling for various ws$_{15}$ lower bounds}
To assess how model performance depends on ws$_{15}$ lower bound several models were trained using the variable sets in rows 5 and 6 in Table \ref{table:setsOfParams}, i.e. $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$, as well as the same set together with DEM. The results can be seen in Table \ref{table:results}. For this investigation comparison was only made with the constant baseline model. In all cases adding the DEM information is seen to increase the model performance. The reason for the difference between the 10 m/s lower bound results seen in Tables \ref{table:setsOfParams} and \ref{table:results} is they were obtained with different training setups.

@@

\section{Results for specific locations}
There are two ways to incorporate location in the model building: To include the station location in the training data, and to train on subsets of the data, e.g. by station or groups of stations. In this section the latter approach is considered. Results of an investigation for a few stations can be seen in Table \ref{table:specific_sites}. For many of the stations, the results are worse than expected, calling for further investigation.

@@

% Færum Mynd 4.1 fram fyrir töflu 4.4.   
% KOMNIR HINGAÐ

Instead of looking at the exact values of MAPE at select stations, a plot of the error distribution can be created. This can be seen in Figure (\ref{fig:errorMap}). Looking at Figure (\ref{fig:errorMap}), the worst performing stations can be seen at Vestfirðir and around the coastline. Stations further inland seem to have lower error. The worst performing station, at Seljalandsdalur, is at Vestfirðir while the best performing station Garðskagaviti is at the South-Western tip of Iceland in Reykjanes.

\begin{figure}[h]
    \centering
    \includegraphics[scale = 0.5]{Figures/errorMap.png}
    \caption[MAPE error distribution of stations shown on a map of Iceland.]{The MAPE error of each station in data shown as color gradient circles. That is each station is represented by one circle, with the error value represented as a color gradient from dark blue to yellow. The lowest error at a single station was around 7\% at Garðskagaviti and the highest around 35\% at Seljalandsdalur. It is important to note that the model is trained using a cutoff of 10 m/s and this cutoff point is determined using the reanalysis wind at 15 meters above ground.}
    \label{fig:errorMap}
\end{figure}

@@

Table (\ref{table:closed_intervals}) shows no improvement over the values in the last column in Table (\ref{table:results}). As previously mentioned, the gust factor decreases with increasing wind speed and thus, training on intervals and lessening this effect might be expected to give better results. This does not seem to be the case. Some interesting sites to look closer at for drivers might include places such as Ingólfsfjall, Kjalarnes and others. These can be seen in Table (\ref{table:more_specific_sites}).

>>>>

The resulting model will be referred to as the \textit{primary model} and the corresponding parameters as the \textit{primary parameter set}. The .

\begin{table}[h]
  \caption[Model results for different sets of parameters.]{Models defined by a selection of parameter sets, with increasing complexity. The table shows decreasing MAPE as more variables are added.}
    \label{table:setsOfParams}
    \centering
    \begin{tabular}{lc}
        \toprule
        Model variables & MAPE\\
        \midrule
        $[\,]$ Baseline & 23.9\%\\
        $[ws_{15}]$ Baseline & 17.2\%\\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}]$ & 16.7\% \\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd]$ & 13.8\% \\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ & 13.7\%\\
        $[ws_{15}, t_{15}, p_{15}, wd_{15}, \text{altitude}, twd, N^2, Ri]$ + DEM* & 12.0\%\\
        $[ws_{15,250,500}, t_{15,250,500}, p_{15,250,500}, wd_{15,250,500}, \text{altitude}, twd, N^2, Ri]$ & 12.0\% \\
        \bottomrule
        *\small{Primary model}
    \end{tabular}
\end{table}

\section{Modeling for various ws$_\textbf{15}$ lower bounds}
To assess how model performance depends on ws$_{15}$ lower bound several models were trained using the variable sets in rows 5 and 6 in Table \ref{table:setsOfParams}, i.e. ws$_{15}$, $t_{15}$, $p_{15}$, wd$_{15}$, altitude, twd, $N^2$, Ri, as well as the same set together with DEM (the primary model). The results can be seen in Table \ref{table:results}. For this investigation comparison was only made with the constant baseline model. In all cases adding the DEM information is seen to increase the model performance. The reason for the difference between the 10 m/s lower bound results seen in Tables \ref{table:setsOfParams} and \ref{table:results} is they were obtained with different training setups.

@@

\section{Results for specific locations}
In this section, the distribution of MAPE by station will be considered. For each station the primary model MAPE was computed using just the ws$_{15} \geq 10$ data for that station was calculated and the results are shown in Figure (\ref{fig:errorMap}). The stations with low performance are all near mountains and often in fjords. The worst cases are in Vestfirðir and Austfirðir, and several bad stations are e.g. in Tröllaskagi and Snæfellsnes, and near Öræfajökull, Esja, Eyjafjallajökull. Inland stations seem to have lower error. The worst performing station is Seljalandsdalur in Vestfirðir while the best performing station is Garðskagaviti at the tip of the Reykjanes peninsula. Evidently rugged landscape makes it difficult to estimate the gust factor accurately.

\begin{figure}[h]
    \centering
    \includegraphics[scale = 0.5]{Figures/errorMap.png}
\caption[MAPE distribution by station]{Mean absolute percentage error (MAPE) at each station, shown as colored circles (dark blue = low error; yellow = high error). The lowest station error is about 7\% at Garðskagaviti and the highest about 35\% at Seljalandsdalur. The model uses a 10 m/s lower bound based on ws$_{15}$.}
\label{fig:errorMap}
\end{figure}

After examining the MAPE at each station individually, it is logical to try accounting for the station (or its location) in the model’s design. There are two ways to do this: To include the station location in the training data, and to train on subsets of the data, e.g. by station or groups of stations. Here the latter approach is considered. Results of an investigation for a few stations can be seen in Table \ref{table:specific_sites}. For many of the stations, the results are worse than expected, sometimes much worse, calling for further investigation.

@@

\section{Results for wind speed intervals}
In addition to training for individual stations one can also try to train for individual wind speed intervals. Table \ref{table:closed_intervals} shows the results of training the primary model

@@

Table (\ref{table:closed_intervals}) shows no improvement over the values in the last column in Table (\ref{table:results}). As previously mentioned, the gust factor decreases with increasing wind speed and thus, training on intervals and lessening this effect might be expected to give better results. This does not seem to be the case. Some interesting sites to look closer at for drivers might include places such as Ingólfsfjall, Kjalarnes and others. These can be seen in Table \ref{table:more_specific_sites}.
